\chapter{Discrete and Masked Diffusion Language Models}
\label{ch:discrete-diffusion}

This chapter moves from the continuous denoising picture to the discrete
language setting. The state is a token sequence, corruption is categorical, and
the reverse model predicts tokens rather than vectors. The purpose of the
chapter is to define the base masked diffusion mechanics used throughout the
thesis. Corrector timing is deferred to Chapters~\ref{ch:informed-correctors}
and~\ref{ch:contribution}.

\section{Discrete State Spaces}
\label{sec:ch3-discrete-state}

Let $\mathcal{V}$ be a finite vocabulary and let $L$ be the sequence length.
A clean text sequence is
\[
  x = (x^1,\ldots,x^L) \in \mathcal{V}^L .
\]
A discrete diffusion process defines a corruption process on this finite state
space or on an enlarged state space. Unlike the Gaussian case, corruption
cannot be expressed as adding a small real-valued perturbation. Instead, tokens
are replaced, masked, or moved through a continuous-time Markov chain on
categorical states \parencite{lou2024sedd}.

One general formulation uses a rate matrix on the vocabulary. If $p_t$ denotes
the row vector of probabilities at time $t$, then a continuous-time Markov
chain evolves according to the Kolmogorov forward equation
\[
  \frac{\dif}{\dif t}p_t = p_t Q,
\]
where $Q$ is a rate matrix. Different choices of $Q$ produce different
corruption processes. The absorbing-mask process used in masked diffusion is a
particularly simple and useful case.

\section{Absorbing-Mask Corruption}
\label{sec:ch3-absorbing-mask}

Masked diffusion introduces a special token $\Mask$ and extends the state
space to $\mathcal{V}\cup\{\Mask\}$. In the forward process, each clean token
is independently replaced by $\Mask$ with a time-dependent probability. If
$z_t$ is the corrupted sequence at time $t$, then a typical absorbing-mask
transition has the form
\[
  q(z_t \mid x)
  = \prod_{\ell=1}^L q(z_t^\ell \mid x^\ell),
\]
where $z_t^\ell$ is either the original token $x^\ell$ or the absorbing mask
token. Once a position has been masked in the forward process, it remains
masked as the noising time increases.

The reverse process starts from a heavily masked sequence and progressively
fills in tokens. This resembles masked language modelling, but the use is
different. Masked language modelling trains a denoiser on randomly corrupted
examples. Masked diffusion uses the denoiser repeatedly as part of a generative
trajectory.

\section{Reverse Denoising and the Predictor}
\label{sec:ch3-reverse-predictor}

The learned reverse model is a predictor. Given a partially masked state
$z_t$, it outputs token probabilities for positions that may be filled or
updated. In an MDLM-style model, the network directly parameterises
$p_\theta(x \mid z_t)$ or a closely related distribution over clean tokens
\parencite{sahoo2024mdlm}. Generation proceeds by choosing some masked
positions and sampling or selecting tokens from these probabilities.

Let $T$ be the denoising horizon. A sample path is a sequence
\[
  z_T,z_{T-1},\ldots,z_0,
\]
where $z_T$ is highly masked and $z_0$ is a completed token sequence. The
predictor schedule determines how many positions are unmasked at each step and
which positions are eligible. A common principle is easy-first decoding:
positions with high confidence are committed earlier.

This predictor schedule is not the object optimised in the thesis. It is part
of the fixed backend. The thesis studies extra corrector placements on top of
this fixed trajectory.

\section{Committed Tokens and Revisable Sets}
\label{sec:ch3-revisable-sets}

At any step some positions remain masked and some have already been assigned
tokens. The sampler must specify which positions are revisable. For a pure
unmasking sampler, committed positions may never be changed. For a
self-correction backend, committed positions may remain eligible for later
revision. We denote the revisable or action set at step $t$ by $R_t$.

The set $R_t$ is a backend property. It is not the same as the set of masked
positions unless the backend defines it that way. This distinction becomes
important for informed correctors: signals such as entropy, inverse margin, or
quality mass are only meaningful as timing signals if they are computed on the
same set of positions the corrector can actually act on.

\section{Factorisation and Approximation Error}
\label{sec:ch3-factorisation}

The exact reverse conditional of a discrete diffusion model is generally
intractable because it involves dependencies among many token positions. Scaled
models therefore use factorised approximations, predicting or sampling
positions through products of local categorical distributions. This is
computationally necessary but introduces approximation error.

\textcite{lavenant2025lz} analyse masked diffusions with factorised
approximations and separate sources of error into learning error and
factorisation error. The learning error is controlled by the quality of the
trained denoiser. The factorisation error is tied to the inference procedure:
how the reverse process commits to tokens, how schedules are chosen, and
whether additional correction steps can reduce the approximation gap. This
error-decomposition viewpoint motivates the later focus on inference-time
correction.

\section{Base Model Context}
\label{sec:ch3-base-model-context}

Several recent model families instantiate discrete or masked diffusion for
language. MDLM provides a simple masked diffusion language-model formulation
with an effective training objective \parencite{sahoo2024mdlm}. SEDD develops
discrete diffusion through ratio estimation and a score-entropy objective
\parencite{lou2024sedd}. LLaDA and Dream scale diffusion-style language models
toward large language-model regimes \parencite{nie2025llada,ye2025dream}.

For this thesis, these models supply background rather than alternative thesis
objects. The theory in Chapter~\ref{ch:contribution} does not assume a
particular base model. It assumes only that, after fixing a predictor and
corrector, one can define and evaluate the gain of applying the corrector at a
set of time steps.

\section{Common Random Numbers}
\label{sec:ch3-crn}

Discrete generation involves random choices: initial masks, token sampling,
tie-breaking, and possibly corrector transitions. When comparing two schedules,
one should hold all irrelevant randomness fixed. This is the common-random-
number principle. A baseline trajectory and a corrected trajectory should share
the same seed and predictor randomness, differing only in the corrector
intervention being evaluated.

This principle is methodological rather than model-specific. It is what makes
the schedule gain $G(S)$ in Chapter~\ref{ch:contribution} interpretable as the
effect of a schedule rather than noise from unrelated sampling variation.

\section{Summary}
\label{sec:ch3-summary}

Masked diffusion language models generate by progressively replacing mask
tokens with vocabulary tokens. The base predictor schedule defines the main
trajectory, while the revisable set $R_t$ determines which positions can be
acted on at each step. The next chapter surveys the model families and
inference-time mechanisms around this base process, with careful separation
between remasking, informed refinement, trajectory search, and corrector
timing.
