\chapter{Model Families and Inference-Time Mechanics}
\label{ch:masked-diffusion}

The previous chapter described the mechanics of masked diffusion. This chapter
places the main model families and inference-time methods into a taxonomy. The
taxonomy matters because many methods improve masked diffusion generation, but
they do so by changing different parts of the pipeline. The thesis concerns
fixed-budget timing of a fixed informed corrector. It is therefore necessary to
separate base models, remasking samplers, informed refinement mechanisms, and
trajectory-search methods.

\section{MDLM-Style Masked Diffusion}
\label{sec:ch4-mdlm}

MDLM is the closest base-model ancestor of the empirical backend used in this
thesis \parencite{sahoo2024mdlm}. Its state is a partially masked token
sequence, and its denoiser predicts clean tokens from that state. Sampling
starts from a highly masked sequence and repeatedly unmasks positions until a
complete sequence is produced.

The important feature for this thesis is the separation between the base
predictor and any additional correction. The predictor supplies the default
trajectory. It determines the horizon $T$, the masking or unmasking schedule,
and the intermediate states at which a corrector could be inserted. Once this
predictor is fixed, corrector timing becomes an additional layer.

SEDD offers a different discrete diffusion formulation based on estimating
ratios of the data distribution \parencite{lou2024sedd}. LLaDA and Dream show
that diffusion-style language generation can be scaled to large language-model
settings \parencite{nie2025llada,ye2025dream}. These models are part of the
broader masked/discrete diffusion landscape. They are not themselves a timing
theory.

\section{Remasking and Iterative Refinement Samplers}
\label{sec:ch4-remasking}

Remasking methods change the inference process by allowing tokens to be masked
again or by revisiting uncertain positions through a modified sampler.
ReMDM is the main example in the thesis taxonomy
\parencite{wang2025remdm}. It introduces remasking strategies that can improve
generation by changing how the reverse process handles already-filled tokens.

This is adjacent to corrector timing but not identical to it. A remasking
sampler changes the trajectory itself: positions may be returned to a masked
state, the predictor may see states outside the original unmasking path, and
the NFE accounting is tied to a modified inference procedure. In the thesis
framework, by contrast, the predictor trajectory is fixed and an additional
corrector is placed at selected time steps.

Other uncertainty-driven or entropy-driven decoding methods also belong near
this category when their primary object is token ordering, path choice, or
unmasking policy rather than corrector placement
\parencite{chen2025denoisingentropy,benhamu2025ebsampler,kim2025tokenordering,kim2025klass,hong2025upo,he2025mdpo}.
They are relevant because they show that inference-time mechanics matter, but
they do not answer the fixed-budget corrector-timing question directly.

\section{ProSeCo as the Empirical Backend}
\label{sec:ch4-proseco}

ProSeCo is the empirical backend used in this thesis
\parencite{schiff2026proseco}. It is a self-correcting masked diffusion system:
after tokens have been committed by the base predictor, an annealed refinement
mechanism can revise eligible positions. This makes it a natural backend for
studying corrector timing, because the corrector is meaningful, the denoising
trajectory is explicit, and the extra cost of correction can be counted.

In the ProSeCo-OWT case study, the predictor schedule, corrector kernel, and
evaluation functional are held fixed. The variable is the placement schedule:
the set of time steps at which the corrector is applied. The empirical
question is whether simple signals can choose those steps well, or whether the
value of a schedule depends on interactions among placements.

The thesis does not treat ProSeCo as a theory assumption. It is the primary
case study because it is an informed self-correction backend with available
weights and a traceable experiment path. The mathematical framework in
Chapter~\ref{ch:contribution} applies to any fixed predictor/corrector pair
with a definable schedule gain.

\section{PRISM and Quality-Guided Informed Refinement}
\label{sec:ch4-prism}

PRISM is best classified as informed-corrector-adjacent or informed refinement
\parencite{kim2025prism}. It uses quality or reward-guided information to
improve masked diffusion samples. This makes it relevant to the thesis, but it
must be handled carefully.

There are two distinct uses of PRISM-like ideas. First, PRISM as a
non-separable quality-guided refinement mechanism is a possible informed
refinement backend. The thesis does not currently test this use, and therefore
does not rule it out. Second, a PRISM-style quality-mass statistic can be used
as a per-step separable score. The thesis denotes this signal by $QM_t$
(historical raw files may use \texttt{Q\_t}). When used this way, quality mass
belongs to the marginal-ranker baseline analysed by Theorem A in
Chapter~\ref{ch:contribution}.

This distinction prevents an overclaim. A negative result for tested separable
rankers is not a rejection of PRISM as an informed refinement method. It only
limits PRISM-style quality mass when it is reduced to a separable timing score
inside the tested ranker class.

\section{Trajectory Search and Test-Time Scaling}
\label{sec:ch4-search}

Another family of methods spends additional computation by searching over
trajectories, candidates, or refinements. IterRef uses iterative refinement as
a test-time scaling mechanism \parencite{lee2025iterref}. S$^3$ uses
stratified search for diffusion language models \parencite{bilal2026s3}.
Particle Gibbs methods use particle or MCMC ideas to scale inference-time
exploration \parencite{dang2025particlegibbs}. These methods are important
because they show that additional computation can be directed in structured
ways.

They differ from the main thesis object in two respects. First, they usually
search over candidate trajectories or outputs, not just over the time steps at
which a fixed corrector is inserted. Second, they may use reward, verifier, or
true-feedback information that is unavailable in a deployable open-loop timing
rule. In the thesis experiments, search procedures with true schedule-gain
feedback are interpreted as structural existence results: they can show that
good schedules exist, but not necessarily that an inference-time scheduler has
been learned.

\section{Why Timing Is the Object}
\label{sec:ch4-why-timing}

The methods reviewed in this chapter all modify inference-time behaviour, but
at different layers. Base models define the denoising distribution. Remasking
changes the sampler. Informed refinement changes or augments the correction
mechanism. Search methods branch over candidates. Corrector timing is the
remaining trajectory-level allocation question: if the model already has a
corrector and a fixed budget, where should the corrector be placed?

This is the question that can be studied with schedule gains. It is narrow
enough to admit a clean mathematical framework, yet broad enough to include
different model/corrector pairs. The next chapter defines informed correctors
and the signals used for timing, then motivates the formal schedule-selection
framework of Chapter~\ref{ch:contribution}.
