\begin{abstract}
Masked diffusion language models generate by progressively unmasking token
positions along a denoising trajectory. An informed corrector is an additional
inference-time refinement mechanism that can be inserted at selected steps of
this trajectory. When corrector placements are limited by a budget $B$, timing
becomes a combinatorial schedule-selection problem: which $B$ steps out of $T$
should receive a corrector call?

This thesis develops a model- and corrector-agnostic framework for fixed-budget
informed-corrector timing. The framework defines a schedule gain $G(S)$, a
marginal additive surrogate $A(S)$, pairwise interactions $\xi_{t,t'}$, a
pairwise surrogate $Q(S)$, and a regime-classification protocol. The central
theoretical contributions are Theorem~A, a marginal proxy regret bound
identifying when separable per-step ranking is near-optimal, and
Theorems~B/B$'$, a pairwise surrogate regret framework with a finite-pool
high-probability form suited to experimental validation. Diagnostic
Framework~C classifies a given model/corrector pair into no-op,
marginal/rankable, interaction-driven, higher-order, or online-decision-like
regimes.

The empirical case study is ProSeCo-OWT, a self-correcting masked diffusion
system. Prior baseline results indicate that tested separable rankers do not
recover the Monte Carlo oracle headroom at budgets $B\in\{2,3,4\}$, while
search procedures using true schedule-gain feedback recover 49--84\% of
that headroom. These empirical claims are treated as pending Phase~0
reproducibility re-confirmation before use as final thesis evidence.

The thesis is deliberately scoped: it studies timing of a fixed corrector
under a fixed predictor, not remasking, token selection, corrector-kernel
design, or generic test-time scaling. Limitations include single-backbone
empirical scope, pending re-validation of prior results, and the absence of
a deployable Level-3 feature-conditioned scheduler.
\end{abstract}
\newpage
