\chapter{Introduction}
\label{ch:introduction}

Language generation is usually presented as a left-to-right problem: a model
conditions on a prefix and samples the next token. Diffusion language models
offer a different view. They begin from a corrupted or masked sequence and
generate by repeatedly denoising it. This makes generation a trajectory rather
than a single autoregressive pass. At each step the model has a partially
specified text, a set of uncertain positions, and a choice about how much
inference-time effort to spend before moving on.

This thesis studies one such inference-time choice. We fix a masked diffusion
language model, fix its predictor schedule, fix an informed corrector, and ask
when the corrector should be applied under a fixed budget. The question is not
whether diffusion language models are better than autoregressive models. It is
also not whether a new corrector kernel can be designed. The question is more
specific: given a corrector that can refine a denoising trajectory, where along
that trajectory should a limited number of corrector placements be spent?

\section{Diffusion Models and Language Generation}
\label{sec:ch1-diffusion-language}

Diffusion models learn to reverse a gradual corruption process. In image
generation, the usual corruption adds Gaussian noise and the reverse process
removes it by estimating a score or denoising direction
\parencite{ho2020ddpm,song2021score}. In language, the state space is
discrete. A token is not a point in Euclidean space that can be nudged by a
small Gaussian vector. Masked and discrete diffusion language models therefore
replace continuous noise with categorical corruption, most commonly by masking
tokens and learning to reconstruct them
\parencite{sahoo2024mdlm,lou2024sedd,nie2025llada,ye2025dream}.

This gives masked diffusion language models a useful structural property:
generation unfolds through a sequence of partially completed texts. Early in
the trajectory many positions are masked. Later, more positions have been
committed to concrete tokens. This trajectory exposes intermediate
uncertainty, confidence, and quality information that can be used at inference
time. It also creates several distinct design questions. One may ask how to
train the base denoiser, how to choose the predictor schedule, how to decide
which tokens to unmask, whether to remask tokens, how to design a correction
kernel, or when to apply a fixed correction kernel. This thesis focuses only
on the last of these questions.

\section{Correction as Inference-Time Effort}
\label{sec:ch1-correction-effort}

A predictor step advances the masked diffusion trajectory. A corrector step
spends additional computation at the current state to improve or refine the
sample. The continuous diffusion literature often expresses this as a
predictor-corrector sampler, where a numerical reverse step is followed by an
MCMC-like refinement step \parencite{song2021score}. In discrete diffusion,
correctors are typically Markov kernels or refinement mechanisms that use
model-derived information to revise tokens or local states
\parencite{zhao2024informedcorrectors,gat2024dfm,schiff2026proseco}.

The existence of a corrector does not answer the timing question. If applying
the corrector after every denoising step is too expensive, then the model has a
placement budget. With a horizon of $T$ denoising steps and a budget of $B$
corrector placements, timing becomes a schedule-selection problem: choose
$B$ steps out of $T$. This problem is combinatorial even before one considers
adaptive or feature-conditioned policies.

The timing question is under-specified in much of the surrounding literature.
Remasking methods change the sampler by returning positions to a masked or
uncertain state \parencite{wang2025remdm}. Token-ordering methods study which
positions should be committed first \parencite{kim2025tokenordering}. Search
and test-time scaling methods branch or refine candidate trajectories using
external scores or rewards \parencite{lee2025iterref,bilal2026s3}. These are
important adjacent directions, but they do not by themselves define a theory
of fixed-budget corrector placement for a fixed predictor and corrector.

\section{Main Research Question}
\label{sec:ch1-research-question}

The central question of the thesis is:
\begin{quote}
For a fixed predictor schedule and fixed corrector-placement budget $B$ in a
masked diffusion language model, when is informed correction timing reducible
to marginal signal ranking, and when does it require interaction-aware or
search-based scheduling?
\end{quote}

Here $B$ is a placement budget, not merely a generic number-of-function-
evaluations budget. If each placement costs $c_{\mathrm{corr}}$ additional
forward passes, then the associated extra-NFE budget is
$B_{\mathrm{NFE}} = c_{\mathrm{corr}}B$. This distinction matters because the
mathematical object is the set of time steps at which the corrector is placed.

The thesis develops a model- and corrector-agnostic framework. For any fixed
masked diffusion predictor, informed corrector, quality functional $F$, and
placement budget $B$, the framework defines the schedule gain $G(S)$, marginal
gain $\Delta_t$, additive surrogate $A(S)$, pairwise interaction
$\xi_{t,t'}$, pairwise surrogate $Q(S)$, and diagnostic quantities that
classify the timing regime. ProSeCo-OWT is the primary empirical case study,
not an assumption of the theory \parencite{schiff2026proseco}.

\section{What This Thesis Is Not About}
\label{sec:ch1-non-goals}

The scope is deliberately narrow.

First, this thesis does not optimise the predictor schedule. The predictor
schedule determines how the base masked diffusion sampler moves from a masked
sequence to a completed sequence. In this work that schedule is fixed.

Second, the thesis does not study token-selection policy as the primary
object. A corrector may choose positions inside a revisable set, but the thesis
asks when the corrector is called, not which token is revised once it is called.

Third, the thesis does not present a new remasking sampler. Remasking can be an
effective inference-time strategy, but it changes the state dynamics rather
than allocating a fixed corrector kernel along a fixed trajectory.

Fourth, the thesis does not design a new corrector kernel. Corrector kernels
are reviewed because they motivate the problem, but the contribution is at the
trajectory level: fixed kernel, fixed budget, variable timing.

\section{Contributions}
\label{sec:ch1-contributions}

The thesis makes five contributions.

\begin{enumerate}
  \item It formulates fixed-budget informed-corrector timing as a schedule-gain
  problem. The formulation is model- and corrector-agnostic and separates
  timing from remasking, token selection, predictor scheduling, and kernel
  design.

  \item It proves Theorem A, a marginal-ranker baseline. The theorem states
  when a separable proxy score is sufficient: if schedule gains are
  approximately additive and the proxy is uniformly calibrated to the marginal
  gains, top-$B$ ranking has bounded regret.

  \item It develops Theorem B and Theorem B$'$, a pairwise-surrogate
  framework. These results identify when a pairwise approximation to
  schedule gain is useful, including an experimentally usable finite-pool
  form with a no-leakage requirement.

  \item It introduces Diagnostic Framework C. This protocol classifies a fixed
  empirical tuple into no-op, marginal/rankable, interaction-driven,
  higher-order/chaotic, or online-decision-like regimes.

  \item It uses ProSeCo-OWT as an empirical case study. Prior baseline results
  suggest that tested separable rankers do not recover the observed MC-oracle
  headroom, while search procedures using true schedule-gain feedback recover
  a substantial fraction of that headroom. These empirical claims are treated
  as pending Phase 0 re-confirmation before being used as final thesis
  evidence.
\end{enumerate}

The contribution is therefore not a universal impossibility claim about
marginal signals or ranking. The negative statement is narrower: for the
tested separable rankers on ProSeCo-OWT, prior results do not recover the
MC-oracle headroom. The theory leaves room for feature-conditioned rankers,
pairwise schedulers, online controllers, non-separable PRISM-style refinement,
and search procedures.

\section{Scope and Limitations}
\label{sec:ch1-limitations}

The main empirical risk is external validity. ProSeCo-OWT is a useful and
audited case study, but it is one backend. A bounded LLaDA-SFT probe did not
establish transferable MC-oracle headroom at the tested resolution. The thesis
therefore avoids claiming that the observed ProSeCo-OWT regime is universal.

A second limitation is deployability. Search procedures such as coordinate
descent with true gain feedback are structural existence results: they show
that better schedules exist under the measured quality functional, but they do
not by themselves give an inference-time scheduler. A deployable scheduler
claim requires a feature-conditioned rule evaluated on held-out seeds, as
formalised in Chapter~\ref{ch:contribution}.

A third limitation is metric dependence. The quality functional $F$ used in
the ProSeCo-OWT case study is a relative evaluation metric. The theory accepts
any fixed $F$, but empirical conclusions are always conclusions about the
chosen metric.

\section{Thesis Roadmap}
\label{sec:ch1-roadmap}

Chapter~\ref{ch:background-diffusion} reviews diffusion models as iterative
denoising procedures and introduces the Markov-chain viewpoint used later.
Chapter~\ref{ch:discrete-diffusion} moves to discrete and masked diffusion,
where categorical corruption and absorbing masks replace Gaussian noise.
Chapter~\ref{ch:masked-diffusion} surveys masked diffusion language-model
families and inference-time mechanics, carefully separating base models,
remasking, informed refinement, and trajectory search. Chapter~\ref{ch:informed-correctors}
defines informed correctors and builds the conceptual bridge from corrector
kernels to corrector timing. Chapter~\ref{ch:contribution} gives the formal
schedule-gain framework, theorems, diagnostics, and limitations. The remaining
chapters report experiments and conclusions.
