\chapter{Background: Continuous Diffusion Models}
\label{ch:background-diffusion}

Diffusion models define a generative process by learning to reverse a gradual noising
procedure. In the continuous setting, this noising procedure is a stochastic differential
equation (SDE) that transforms structured data into unstructured noise; generation then
amounts to integrating the time-reversed SDE. The foundational ideas --- score matching,
the reverse SDE, and predictor-corrector sampling --- transfer directly to the discrete
masked setting studied in this thesis, and this chapter develops them in the form that
makes that transfer cleanest.

The chapter is organised as follows. Section~\ref{sec:score-matching} reviews the score
function and the denoising score matching objective. Section~\ref{sec:diffusion-sde}
develops the SDE framework and places DDPM within it.
Section~\ref{sec:diffusion-sampling} covers the main families of samplers, with
particular attention to the predictor-corrector scheme whose discrete analogue is the
subject of this thesis. Section~\ref{sec:cfg} briefly covers classifier-free guidance.
Section~\ref{sec:ch2-summary} concludes by mapping the continuous concepts onto their
discrete counterparts.

%% ---------------------------------------------------------------
\section{Score Matching and the Denoising Objective}
\label{sec:score-matching}

\subsection{The score function}
\label{ssec:score-function}

Let $p_t$ denote the marginal distribution of the noisy data at time $t \in [0, 1]$,
with $p_0 = \pdata$ the data distribution and $p_1$ a tractable noise distribution (in
practice, an isotropic Gaussian). The \emph{score function} of $p_t$ is
\begin{equation}
  \nabla_{x} \log p_t(x),
  \label{eq:score-function}
\end{equation}
which points in the direction of increasing log-probability. If one had access to the
true score at every noise level, one could integrate the reverse SDE and recover samples
from $\pdata$ (Anderson~\citeyear{anderson1982reverse}; see
Section~\ref{sec:diffusion-sde}). The core idea of score-based generative modelling is
to train a neural network $\stheta(x, t)$ to approximate this quantity.

\subsection{Score matching}
\label{ssec:score-matching}

\textcite{song2019ncsn} propose learning the score by minimising the \emph{score
matching} objective \parencite{hyvarinen2005estimation},
\begin{equation}
  \mathcal{L}_{\mathrm{SM}}(\theta)
  = \E_{t}\E_{p_t(x)}\!\left[
      \tfrac{1}{2}\norm{\stheta(x,t)}^2 + \nabla_x \cdot \stheta(x,t)
    \right].
  \label{eq:score-matching}
\end{equation}
The divergence term makes this objective impractical at high dimension. The key insight of
\textcite{vincent2011connection} is that the score of the marginal can be estimated via
the score of the tractable transition kernel:
\begin{equation}
  \nabla_{x_t} \log p_t(x_t)
  = \E\!\left[\nabla_{x_t} \log q(x_t \mid x_0) \;\Big|\; x_t\right].
  \label{eq:score-posterior}
\end{equation}
This yields the \emph{denoising score matching} (DSM) objective,
\begin{equation}
  \mathcal{L}_{\mathrm{DSM}}(\theta)
  = \E_{t}\E_{x_0 \sim \pdata}\E_{x_t \sim q(\cdot\mid x_0)}\!\left[
      \norm{\stheta(x_t, t) - \nabla_{x_t} \log q(x_t \mid x_0)}^2
    \right],
  \label{eq:dsm}
\end{equation}
where $q(x_t \mid x_0)$ is the forward transition at time $t$ given the clean datum $x_0$.
The DSM loss~\eqref{eq:dsm} has the same minimiser as the intractable
objective~\eqref{eq:score-matching} (up to a constant), but is computable by simulation
from the forward process. Training thus requires only forward-process samples, not
score evaluations.

%% ---------------------------------------------------------------
\section{Diffusion as a Stochastic Process}
\label{sec:diffusion-sde}

\subsection{The forward SDE}
\label{ssec:forward-sde}

A broad family of diffusion models can be unified under the framework of
\textcite{song2021score}, who describe the forward process as the It\^o SDE
\begin{equation}
  \dif x_t = f(x_t, t)\,\dif t + g(t)\,\dif \Wt,
  \label{eq:forward-sde}
\end{equation}
where $f(\cdot,t) : \mathbb{R}^d \to \mathbb{R}^d$ is the \emph{drift}, $g(t) > 0$ is
the scalar \emph{diffusion coefficient}, and $\Wt$ is a standard Brownian motion. The
two canonical choices of drift and diffusion give:
\begin{itemize}[nosep]
  \item \textbf{Variance-preserving (VP-SDE)}: $f(x,t) = -\tfrac{1}{2}\beta(t) x$,
        $g(t) = \sqrt{\beta(t)}$, which preserves $\E[\norm{x_t}^2]$ and sends $p_t$
        toward $\mathcal{N}(0, I)$.
  \item \textbf{Variance-exploding (VE-SDE)}: $f = 0$,
        $g(t) = \sqrt{\dif \sigma^2(t)/\dif t}$, which inflates noise variance without
        normalising the signal.
\end{itemize}
In both cases the marginal $p_t(x_t \mid x_0)$ is Gaussian with computable mean and
variance, making simulation from the forward process straightforward.

\subsection{DDPM as a discrete-time special case}
\label{ssec:ddpm}

\textcite{ho2020ddpm} introduce the Denoising Diffusion Probabilistic Model (DDPM) in a
discrete-time formulation, which maps cleanly onto the VP-SDE with a specific linear
noise schedule. Define a sequence of variances $0 < \beta_1 < \cdots < \beta_T \ll 1$
and set $\abardpm_t \coloneq \prod_{s=1}^{t}(1 - \betanoise_s)$. The forward process is
\begin{equation}
  q(x_t \mid x_0) = \mathcal{N}(x_t;\; \sqrt{\abardpm_t}\, x_0,\;
    (1 - \abardpm_t) I),
  \label{eq:ddpm-forward}
\end{equation}
so $x_t = \sqrt{\abardpm_t}\,x_0 + \sqrt{1-\abardpm_t}\,\varepsilon$,
$\varepsilon \sim \mathcal{N}(0,I)$, giving a reparametrised forward sample in a single
step. As $\abardpm_t \to 0$, the signal is fully destroyed and $x_T \approx
\mathcal{N}(0, I)$.

The DDPM training objective is equivalent to DSM: the network $\etheta(x_t, t)$
predicts the noise $\varepsilon$ added to $x_0$, and training minimises
\begin{equation}
  \mathcal{L}_{\mathrm{DDPM}}(\theta)
  = \E_{t, x_0, \varepsilon}\!\left[
      \norm{\etheta(\sqrt{\abardpm_t}\,x_0 + \sqrt{1-\abardpm_t}\,\varepsilon,\; t)
            - \varepsilon}^2
    \right].
  \label{eq:ddpm-loss}
\end{equation}
The score can be recovered from the noise prediction via
$\stheta(x_t, t) = -\etheta(x_t, t) / \sqrt{1 - \abardpm_t}$. An equivalent
parameterisation predicts $\xhat_\theta(x_t, t) \approx x_0$ directly, which will be the
natural analogue in the masked discrete setting.

\subsection{The reverse SDE}
\label{ssec:reverse-sde}

The key theoretical result enabling score-based generation is
\textcite{anderson1982reverse}: a diffusion process run forward according
to~\eqref{eq:forward-sde} has a time-reversed counterpart that satisfies
\begin{equation}
  \dif x_t = \left[f(x_t, t) - g(t)^2 \nabla_{x_t} \log p_t(x_t)\right]\dif t
             + g(t)\,\dif \bar{W}_t,
  \label{eq:reverse-sde}
\end{equation}
where $\bar{W}_t$ is a Brownian motion running backward in time. Replacing the
intractable score $\nabla_{x_t} \log p_t(x_t)$ with the learned $\stheta(x_t, t)$ and
discretising yields the reverse sampling chain. The generation procedure is therefore:
(i)~sample $x_T \sim p_T \approx \mathcal{N}(0, I)$; (ii)~integrate the reverse
SDE~\eqref{eq:reverse-sde} from $t = T$ to $t = 0$.

%% ---------------------------------------------------------------
\section{Sampling from Diffusion Models}
\label{sec:diffusion-sampling}

\subsection{Euler--Maruyama discretisation}
\label{ssec:euler-maruyama}

The straightforward approach is to discretise the reverse SDE via the Euler--Maruyama
scheme with stepsize $\Delta t$:
\begin{equation}
  x_{t - \Delta t} = x_t
    - \left[f(x_t, t) - g(t)^2 \stheta(x_t, t)\right]\Delta t
    + g(t)\sqrt{\Delta t}\,z_t,
  \quad z_t \sim \mathcal{N}(0,I).
  \label{eq:euler-maruyama}
\end{equation}
This yields the ancestral sampling procedure of \textcite{ho2020ddpm}. The discretisation
error decreases with $\Delta t$, so reducing the step count (to save model evaluations)
increases the gap between the generated distribution and the true reverse process.

\subsection{Deterministic ODE samplers and DDIM}
\label{ssec:ddim}

Every diffusion model also admits a deterministic sampling trajectory via a probability
flow ODE \parencite{song2021score}: the ODE whose marginals match those of the forward
SDE is
\begin{equation}
  \frac{\dif x_t}{\dif t}
  = f(x_t, t) - \tfrac{1}{2}g(t)^2 \nabla_{x_t}\log p_t(x_t).
  \label{eq:probability-flow-ode}
\end{equation}
\textcite{song2020ddim} derive \emph{Denoising Diffusion Implicit Models} (DDIM) by
directly constructing a non-Markovian inference process whose marginals agree with
DDPM's but whose reverse step is deterministic. This allows high-quality generation in
far fewer function evaluations (e.g.\ 50 steps instead of 1000) at the cost of slightly
reduced sample diversity. The practical tradeoff between step count, sample quality, and
diversity motivates the \emph{number of function evaluations} (NFE) budget framework
that will be central in Chapter~\ref{ch:masked-diffusion}.

\subsection{Predictor-corrector samplers}
\label{ssec:predictor-corrector}

\textcite{song2021score} propose \emph{predictor-corrector} (PC) samplers, which
interleave:
\begin{enumerate}[nosep]
  \item a \textbf{predictor} step: a discretised reverse SDE or ODE step that advances
        $x_t$ to $x_{t-\Delta t}$;
  \item a \textbf{corrector} step: a Markov chain Monte Carlo (MCMC) step that leaves
        the current marginal $p_{t-\Delta t}$ invariant and uses additional model
        evaluations to improve the sample.
\end{enumerate}
In the continuous setting the corrector is typically a \emph{Langevin dynamics} step,
\begin{equation}
  x \leftarrow x + \epsilon\,\stheta(x, t) + \sqrt{2\epsilon}\,z,
  \quad z \sim \mathcal{N}(0,I),
  \label{eq:langevin}
\end{equation}
which is the Euler--Maruyama discretisation of the Langevin SDE and is reversible with
respect to $p_t$.

\paragraph{The corrector idea, and its discrete analogue.}
The function of a corrector is to \emph{refine} a sample after the predictor has
committed to an approximate location, spending additional NFEs to reduce the gap between
the current sample distribution and the target marginal. In the continuous case the
refining kernel is Langevin dynamics; in the discrete masked case, where the state space
is a finite vocabulary and there is no gradient to follow, the natural analogue is a
Gibbs-style resampling step --- the targeted equivalent of Langevin dynamics on a
discrete space. This is precisely the setting of Chapters~\ref{ch:masked-diffusion}
and~\ref{ch:informed-correctors}: what is the best Gibbs-style corrector for a masked
diffusion model, and can its design be formally justified?

%% ---------------------------------------------------------------
\section{Conditional Generation and Classifier-Free Guidance}
\label{sec:cfg}

Conditioning a diffusion model on a label or prompt $y$ requires approximating the
conditional score $\nabla_{x_t}\log p_t(x_t \mid y)$, which by Bayes' rule equals
$\nabla_{x_t}\log p_t(x_t) + \nabla_{x_t}\log p_t(y \mid x_t)$. The second term is the
score of a classifier trained on noisy data, which \textcite{dhariwal2021diffusionbeats}
use to guide an unconditional model.

\textcite{ho2022cfg} observe that training two models --- one conditional, one
unconditional --- and linearly interpolating their scores avoids the noisy classifier
entirely:
\begin{equation}
  \nabla_{x_t}\log p_t^{\mathrm{guided}}(x_t \mid y)
  = \nabla_{x_t}\log p_t(x_t)
    + \omega\!\left[\nabla_{x_t}\log p_t(x_t \mid y)
                  - \nabla_{x_t}\log p_t(x_t)\right],
  \label{eq:cfg}
\end{equation}
where $\omega \geq 1$ is the guidance scale. Classifier-free guidance (CFG) is the
standard approach for text-to-image and, increasingly, for language generation.
We note it here for completeness; the informed corrector framework studied in this thesis
does not rely on CFG.

%% ---------------------------------------------------------------
\section{Summary: Continuous Concepts and Their Discrete Counterparts}
\label{sec:ch2-summary}

The continuous diffusion framework rests on four pillars, each of which has a discrete
analogue in the masked diffusion setting:

\begin{center}
\begin{tabular}{@{}lll@{}}
  \toprule
  Concept & Continuous diffusion & Masked diffusion \\
  \midrule
  Noising process & Gaussian SDE & Position-wise Bernoulli masking \\
  Score / denoiser & $\stheta(x_t,t) \approx \nabla \log p_t$ &
    $p_\theta(x_0 \mid z_t)$, token probabilities \\
  Training objective & Denoising score matching & Masked language modelling loss \\
  Corrector & Langevin dynamics & Gibbs resampling of masked positions \\
  \bottomrule
\end{tabular}
\end{center}

\noindent The critical structural difference is the \emph{factorisation approximation}:
in the discrete case, the joint reverse conditional $p(z_{t-1} \mid z_t)$ is intractable,
and one approximates it by a product over positions. This approximation has no continuous
analogue (the Gaussian case is exact), and it is the primary source of sample-quality
degradation in masked diffusion. The error it introduces --- the \emph{factorisation
error} $\Efact$ --- and the question of whether informed correctors can reduce it below
the best unmasking-only bound are the central concerns of this thesis.
