\chapter{Informed Correctors and the Timing Problem}
\label{ch:informed-correctors}

This chapter bridges the literature on informed correctors to the schedule
framework in Chapter~\ref{ch:contribution}. A corrector is an additional
inference-time transition that refines the current state of a diffusion
sampler. An informed corrector uses information from the model, the current
state, a learned quality score, a reward, or another diagnostic to make that
transition more targeted than an uninformed local move.

The thesis does not design a new corrector kernel. It fixes a corrector and
studies when to apply it. This chapter explains why that timing problem is
natural.

\section{Corrector Kernels}
\label{sec:ch5-corrector-kernels}

In the continuous predictor-corrector framework, the predictor advances the
reverse diffusion process and the corrector performs an MCMC-like refinement
step \parencite{song2021score}. In discrete diffusion, the same idea becomes a
local Markov transition on token states. The transition may resample one token,
a block of tokens, or a set of eligible positions. If the transition is
designed to preserve or improve an appropriate target distribution, it can
reduce sampling error at the cost of additional model evaluations.

\textcite{zhao2024informedcorrectors} study informed correctors for discrete
diffusion models, including locally balanced proposals and confidence-guided
selection. The statistical MCMC background connects to locally informed
proposals in discrete spaces \parencite{zanella2020informedproposals} and to
questions about Gibbs or Metropolis-within-Gibbs convergence
\parencite{roberts1997updating,ascolani2024entropycontraction,secchi2025mwggap}.
Discrete Flow Matching, entropy-adaptive Gibbs sampling, Psi-samplers, and
Feynman-Kac correctors provide additional perspectives on correction and
sampling in discrete generative models
\parencite{gat2024dfm,koh2024eags,deschenaux2026psisamplers,hasan2026fkc}.

These works motivate the existence of useful corrector kernels. They do not,
by themselves, solve the placement problem. Once a kernel exists and has a
nontrivial cost, a second question appears: at which denoising steps should it
be used?

\section{Action Sets and Corrective Effort}
\label{sec:ch5-action-sets}

A corrector does not act on an abstract trajectory; it acts on positions in
the current state. Let $R_t$ denote the revisable or action set at step $t$.
This is the set of positions the corrector is allowed to modify. In a
self-correction backend such as ProSeCo, $R_t$ may consist of already
committed positions eligible for revision \parencite{schiff2026proseco}. In
another backend it could be defined differently.

The action set distinguishes correctors from remasking. A remasking sampler
changes the state by returning positions to a masked or uncertain status. A
corrector, as used in this thesis, is a fixed refinement mechanism with a
declared action set. The timing question assumes that this mechanism is already
specified.

Corrective effort is costly. If one corrector placement performs
$c_{\mathrm{corr}}$ additional forward passes, then a schedule with $B$
placements has extra-NFE cost
\[
  B_{\mathrm{NFE}} = c_{\mathrm{corr}}B .
\]
The budget $B$ counts placements; the quantity $B_{\mathrm{NFE}}$ records the
associated compute. Chapter~\ref{ch:contribution} formalises this distinction.

\section{Signals for Timing}
\label{sec:ch5-signals}

An informed timing rule needs signals. The simplest signals are computed at
each denoising step and then used to rank steps. This thesis uses three
representative step signals:
\[
  H_t,\qquad M_t^{-1},\qquad QM_t .
\]
The entropy signal $H_t$ measures aggregate uncertainty over the revisable
set. The inverse margin $M_t^{-1}$ measures lack of confidence by comparing
the top token probabilities. The quality-mass signal $QM_t$ is inspired by
quality-guided or PRISM-style refinement \parencite{kim2025prism}. Historical
raw files may call this signal \texttt{Q\_t}; the thesis writes $QM_t$ to avoid
confusion with the pairwise schedule surrogate $Q(S)$ in
Chapter~\ref{ch:contribution}.

The signals must be computed on the same action set the corrector uses. If the
corrector can only revise positions in $R_t$, then a signal aggregated over a
different set is not a clean proxy for the value of applying the corrector at
step $t$. This signal/action-set consistency condition is one of the Phase 0
pre-flight checks before new experiments are trusted.

\section{The Marginal Timing Baseline}
\label{sec:ch5-marginal-baseline}

The most natural first timing rule is marginal ranking. Let $\Delta_t$ be the
gain from applying the corrector only at step $t$, relative to the no-corrector
baseline. If the gain from a schedule were exactly additive,
\[
  G(S) = \sum_{t\in S}\Delta_t,
\]
then the optimal size-$B$ schedule would be obtained by choosing the $B$ largest
marginal gains. Since the true $\Delta_t$ values are unavailable at inference
time, one replaces them with an observable score such as $H_t$, $M_t^{-1}$, or
$QM_t$.

This gives the baseline tested by Theorem A in Chapter~\ref{ch:contribution}.
Theorem A does not say that marginal ranking always works. It says that if
schedule gains are approximately additive and the proxy score is calibrated to
the marginal gains, then top-$B$ ranking is near-optimal. The theorem is useful
because it makes the failure modes explicit: poor calibration, non-additivity,
or both.

\section{Why Interactions Matter}
\label{sec:ch5-interactions}

Corrector placements may interact. Applying a corrector at an early step can
change the state encountered later. Two placements may be redundant if they
repair the same kind of error, or complementary if one creates a context in
which the other becomes more valuable. In such cases, knowing the singleton
gains $\Delta_t$ is not enough.

The pairwise interaction between two placements is
\[
  \xi_{t,t'} = G(\{t,t'\}) - \Delta_t - \Delta_{t'} .
\]
If these interactions are small, the marginal baseline may be adequate. If
they are structured, an interaction-aware surrogate may explain schedule value
better than the additive surrogate. If higher-order effects dominate, even
pairwise scheduling may fail. Chapter~\ref{ch:contribution} formalises these
possibilities through Theorem B, Theorem B$'$, and Diagnostic Framework C.

\section{ProSeCo-OWT as the Case Study}
\label{sec:ch5-proseco-case-study}

The empirical case study is ProSeCo-OWT. Prior experiments in this repository
used a ProSeCo annealed-refinement corrector with an MDLM-style OpenWebText
backend and horizon $T=64$. The quality functional was negative GPT-2
negative log-likelihood on a fixed window. Those prior results suggest three
facts: there is MC-oracle headroom over uniform placement at budgets
$B\in\{2,3,4\}$; tested separable rankers do not recover that headroom; and
search procedures using true schedule-gain feedback recover a substantial
fraction of it.

These statements are not treated as fresh final evidence until Phase 0
re-confirms the backend and experiment path. The current tests check
determinism, empty-schedule equivalence, single-correction equivalence, common
random numbers, budget accounting, scoring consistency, corrector action sets,
and signal/action-set consistency. Only after those checks pass should new
experiments or stronger empirical claims be made.

\section{From Correctors to the Chapter 6 Framework}
\label{sec:ch5-to-ch6}

The preceding chapters have separated the layers of the problem. A masked
diffusion language model supplies a denoising trajectory. A corrector kernel
supplies an additional refinement transition. Signals supply possible
step-wise timing information. The unresolved question is how to reason about a
budgeted set of placements.

Chapter~\ref{ch:contribution} gives the formal answer. It defines the
schedule gain $G(S)$, the marginal gains $\Delta_t$, the additive surrogate
$A(S)$, pairwise interactions $\xi_{t,t'}$, the pairwise surrogate $Q(S)$, and
diagnostics for deciding whether a model/corrector pair is marginal,
interaction-driven, higher-order, or online-decision-like. The empirical
chapters then test which regime ProSeCo-OWT occupies.
