\chapter{Theoretical Contribution}
\label{ch:contribution}

This chapter develops the mathematical framework used in the thesis. The object
of study is deliberately narrow: given a fixed masked diffusion predictor, a
fixed informed corrector, a fixed trajectory-level quality functional, and a
fixed budget of corrector placements, when should the corrector be applied?

The chapter is self-contained. It begins with the schedule-selection problem,
defines the schedule gain function, formalises the marginal baseline, proves
the baseline regret theorem, and then develops the pairwise surrogate framework
used to diagnose when marginal ranking is insufficient. The empirical backend
used elsewhere in the thesis is ProSeCo-OWT \parencite{schiff2026proseco}, but
the mathematics below is not ProSeCo-specific. It applies to any fixed
predictor/corrector pair for which the schedule gain function can be defined
and estimated.

\section{Motivation and Scope}
\label{sec:ch6-motivation}

In a masked diffusion language model, generation proceeds through a denoising
trajectory. A predictor progressively turns masked positions into tokens.
An informed corrector is an additional refinement mechanism that can be
inserted into this trajectory. The corrector may use model probabilities,
confidence, quality estimates, rewards, verifier information, or other state
information to decide how to refine the current state. Examples of such
mechanisms are reviewed in Chapter~\ref{ch:informed-correctors}, including
informed correctors for discrete diffusion \parencite{zhao2024informedcorrectors}
and self-correcting masked diffusion models \parencite{schiff2026proseco}.

The question here is not how to design the corrector. The corrector kernel is
held fixed. The question is when to spend a limited number of applications of
that fixed corrector. If a corrector can be applied at only $B$ out of $T$
denoising steps, then a timing rule is a schedule-selection rule: it chooses a
subset of $B$ time steps. This makes the problem combinatorial even when the
underlying denoising trajectory is fixed.

This separation is important. Corrector timing is not the same as predictor
scheduling, which changes when and how many tokens are unmasked. It is not
token selection, which decides which positions to revise within a corrector
call. It is not remasking, which changes the sampler by masking already
committed tokens again. It is not corrector-kernel design, which changes the
transition kernel itself. This thesis studies the trajectory-level allocation
of a fixed corrector-placement budget under a fixed predictor schedule.

The framework is model- and corrector-agnostic: for any fixed predictor,
informed corrector, quality functional $F$, and corrector-placement budget
$B$, it defines a schedule gain $G(S)$, marginal gains $\Delta_t$, an additive
surrogate $A(S)$, pairwise interactions $\xi_{t,t'}$, a pairwise surrogate
$Q(S)$, and regime diagnostics. ProSeCo-OWT is the primary empirical case study,
not an assumption of the theory.

\section{Formal Setup}
\label{sec:ch6-formal-setup}

Fix a denoising horizon $T \in \mathbb{N}$. The denoising steps are indexed by
$t \in \{1,\ldots,T\}$. Throughout the chapter the predictor schedule is fixed:
the predictor decides how the masked state evolves when no additional corrector
is inserted. The informed corrector is also fixed: at any step where it is
applied, it uses its prescribed transition rule and hyperparameters.

Let $\mathcal{Y}$ be the space of final generated sequences. Let
$F:\mathcal{Y}\to\mathbb{R}$ be a trajectory-level quality functional. Larger
$F$ is better. In the ProSeCo-OWT experiments, $F$ is the negative GPT-2
negative log-likelihood on a fixed text window; this is used only as a relative
within-run score, not as an absolute language-quality claim.

\begin{definition}[Corrector schedule]
\label{def:ch6-schedule}
A corrector schedule is a subset
\[
  S \subseteq \{1,\ldots,T\}.
\]
The interpretation is that the corrector is applied exactly at the steps in
$S$. The placement budget is an integer $B$, and feasible schedules satisfy
$|S|=B$. This chapter uses binary placement: at most one corrector placement is
allowed at any step.
\end{definition}

Let $y^\emptyset$ denote the final sequence produced by the baseline trajectory,
with no corrector placements. Let $y^S$ denote the final sequence produced when
the corrector is applied at exactly the steps in $S$. Both $y^\emptyset$ and
$y^S$ are random objects unless all randomness is fixed. The theory below can
be read either pathwise, conditioning on common random numbers, or in
expectation over a paired seed distribution. The level must be stated whenever
an empirical claim is made.

\begin{definition}[Schedule gain]
\label{def:ch6-gain}
The gain of schedule $S$ is
\[
  G(S) := F(y^S)-F(y^\emptyset).
\]
Thus $G(S)>0$ means that applying the corrector at schedule $S$ improves the
quality functional relative to the no-corrector baseline under the same paired
conditions.
\end{definition}

The placement budget $B$ counts corrector calls, not raw forward passes. If one
placement costs $c_{\mathrm{corr}}$ extra model forward passes, then the
corresponding extra-NFE budget is
\[
  B_{\mathrm{NFE}} = c_{\mathrm{corr}} B .
\]
For ProSeCo annealed refinement, $c_{\mathrm{corr}}=2$. The mathematical budget
in this chapter is $B$; $B_{\mathrm{NFE}}$ is the corresponding compute cost.
All schedules with the same $B$ have the same placement cost.

\begin{definition}[Exhaustive oracle]
\label{def:ch6-exhaustive-oracle}
An exhaustive budget-$B$ oracle schedule is any maximiser
\[
  S_B^* \in \argmax_{\substack{S\subseteq \{1,\ldots,T\}\\ |S|=B}} G(S).
\]
This is a theorem-level object. In realistic experiments it is not observable
because the number of schedules is $\binom{T}{B}$.
\end{definition}

Empirical work uses weaker oracles. A best-of-$N$ MC-oracle is the best
schedule found among $N$ randomly sampled schedules from a specified sampling
distribution. A pool oracle is the best schedule inside a fixed candidate pool
$C_B$. These are not the exhaustive oracle.

\begin{definition}[MC-oracle and pool oracle]
\label{def:ch6-pool-oracles}
Let $\mathcal{P}_B$ be a sampling distribution over size-$B$ schedules and let
$S^{(1)},\ldots,S^{(N)} \sim \mathcal{P}_B$. A best-of-$N$ MC-oracle is
\[
  S_B^{\mathrm{MC},N}
  \in \argmax_{1\le j\le N} G(S^{(j)}).
\]
For a fixed candidate pool $C_B \subseteq \{S: |S|=B\}$, the pool oracle is
\[
  S_C^* \in \argmax_{S\in C_B} G(S).
\]
The exhaustive oracle is $S_B^*$. The MC-oracle and pool oracle are observable
only relative to their sampled or fixed pools.
\end{definition}

All empirical comparisons in the thesis are paired. The baseline trajectory
and all scheduled trajectories use common random numbers: the same seed,
tokenizer choices, predictor randomness, and evaluation setup are reused, and
only the corrector schedule is changed. This reduces variance and makes
$G(S)$ a schedule effect rather than a mixture of schedule and seed effects.

\section{Action Sets and Signals}
\label{sec:ch6-action-signals}

At each denoising step the corrector has an action set. This is the set of
positions the corrector is allowed to revise. The exact definition depends on
the backend. In ProSeCo, the revisable set consists of positions that have
already been unmasked and are eligible for self-correction.

\begin{definition}[Revisable/action set]
\label{def:ch6-action-set}
Let $R_t \subseteq \{1,\ldots,L\}$ be the revisable, or action, set at step
$t$, where $L$ is the sequence length. If the corrector is applied at step
$t$, its corrected positions must lie inside $R_t$.
\end{definition}

Let $C_t \subseteq R_t$ denote the corrected positions at step $t$ when a
corrector call is made. The condition $C_t\subseteq R_t$ is a backend
invariant: a corrector must not revise positions outside its declared action
set.

The timing problem uses trajectory signals computed at each step. The active
signals in this thesis are:
\[
  H_t,\qquad M_t^{-1},\qquad QM_t .
\]
Here $H_t$ is aggregate entropy over the revisable set, $M_t^{-1}$ is inverse
confidence margin over the revisable set, and $QM_t$ is a quality-mass signal,
inspired by PRISM-style token-quality scoring \parencite{kim2025prism}.
Historical raw result files may use the key \texttt{Q\_t} for this quality
mass; the thesis writes $QM_t$ to avoid collision with the pairwise surrogate
$Q(S)$ defined later.

\begin{definition}[Signal/action-set consistency]
\label{def:ch6-signal-consistency}
A step signal is action-set consistent if it is computed over the same
revisable set $R_t$ on which the corrector acts. In particular, $H_t$,
$M_t^{-1}$, and $QM_t$ must aggregate only over positions in $R_t$, unless a
different set is explicitly declared and justified.
\end{definition}

This consistency condition is not cosmetic. If a signal is computed over a
different set from the one the corrector can act on, then the signal is not a
valid proxy for the marginal value of applying that corrector at that step.

\section{The Marginal Baseline}
\label{sec:ch6-marginal-baseline}

The simplest hypothesis is that the value of a schedule is approximately the
sum of the values of its individual placements. This leads to a marginal
ranking rule.

\begin{definition}[Marginal gain and additive surrogate]
\label{def:ch6-additive}
The single-step marginal gain at step $t$ is
\[
  \Delta_t := G(\{t\}).
\]
The additive surrogate of a schedule $S$ is
\[
  A(S) := \sum_{t\in S}\Delta_t.
\]
\end{definition}

If $G(S)=A(S)$ exactly, then the schedule-selection problem is simple: choose
the $B$ largest $\Delta_t$. In practice $\Delta_t$ is not known at inference
time, so one uses a separable score.

\begin{definition}[Separable ranker]
\label{def:ch6-separable-ranker}
Let $s_t$ denote the observable features or signals at step $t$. A separable
ranker is a scoring rule of the form
\[
  \psi_t = \psi(s_t),
\]
which scores each step independently. The corresponding top-$B$ schedule is
\[
  \widehat S_B(\psi)
  \in \argmax_{\substack{S\subseteq \{1,\ldots,T\}\\ |S|=B}}
       \sum_{t\in S}\psi_t .
\]
Equivalently, $\widehat S_B(\psi)$ is a set of the $B$ largest proxy scores.
\end{definition}

Theorem~\ref{thm:ch6-a} below formalises when this marginal rule is safe. It is
the baseline theorem. The later pairwise theory is needed precisely when this
baseline is empirically inadequate.

\section{Theorem A: Marginal Proxy Regret}
\label{sec:ch6-theorem-a}

Theorem A states that top-$B$ marginal ranking is near-optimal if two
conditions hold: schedules are approximately additive, and the proxy score
calibrates the single-step marginal gains.

\begin{assumption}[Binary placement]
\label{assum:ch6-a-binary}
Feasible schedules have exactly $B$ binary placements:
\[
  S\subseteq \{1,\ldots,T\},\qquad |S|=B.
\]
\end{assumption}

\begin{assumption}[Approximate additivity]
\label{assum:ch6-a-additivity}
There exists $\eta_B\ge 0$ such that
\[
  \abs{G(S)-A(S)} \le \eta_B
  \qquad\text{for every } S \text{ with } |S|\le B .
\]
\end{assumption}

\begin{assumption}[Proxy calibration]
\label{assum:ch6-a-calibration}
There exists $\varepsilon\ge 0$ such that
\[
  \abs{\Delta_t-\psi_t}\le \varepsilon
  \qquad\text{for every } t\in\{1,\ldots,T\}.
\]
\end{assumption}

\begin{lemma}[Top-$B$ under exact additivity]
\label{lem:ch6-topb-additive}
If $A(S)=\sum_{t\in S}\Delta_t$ and $|S|=B$, then any set of the $B$ largest
$\Delta_t$ maximises $A(S)$ among all size-$B$ schedules.
\end{lemma}

\begin{proof}
Let $S$ be any size-$B$ schedule. If there exist $t\in S$ and
$t'\notin S$ with $\Delta_{t'}>\Delta_t$, define
$S'=(S\setminus\{t\})\cup\{t'\}$. Then
\[
  A(S')-A(S)=\Delta_{t'}-\Delta_t>0.
\]
Thus replacing a selected lower-gain step by an unselected higher-gain step
strictly improves the additive objective. Repeating this exchange until no
such pair remains yields a set containing only $B$ largest marginal gains. At
that point no single exchange can improve $A$, and the resulting set is
globally optimal because $A$ is a sum of independent selected terms.
\end{proof}

\begin{lemma}[Proxy top-$B$ regret]
\label{lem:ch6-proxy-regret}
Suppose $\abs{\Delta_t-\psi_t}\le \varepsilon$ for all $t$. Let
$S_A^*\in\argmax_{|S|=B}A(S)$ and let $\widehat S_B$ be any top-$B$ set under
the proxy scores $\psi_t$. Then
\[
  A(S_A^*)-A(\widehat S_B)\le 2B\varepsilon .
\]
\end{lemma}

\begin{proof}
The schedules $S_A^*$ and $\widehat S_B$ have the same cardinality. Therefore
the two difference sets
\[
  O := S_A^*\setminus \widehat S_B,
  \qquad
  P := \widehat S_B\setminus S_A^*
\]
have the same size, at most $B$. Pair the elements of $O$ and $P$ arbitrarily.
For a pair $(o,p)\in O\times P$, the proxy top-$B$ property implies
$\psi_p\ge \psi_o$: otherwise $p$ would not be selected ahead of $o$ by a
top-$B$ proxy rule. By calibration,
\[
  \Delta_o \le \psi_o+\varepsilon
            \le \psi_p+\varepsilon
            \le \Delta_p+2\varepsilon .
\]
Thus $\Delta_o-\Delta_p\le 2\varepsilon$ for every paired swap. Since common
elements cancel in $A(S_A^*)-A(\widehat S_B)$,
\[
  A(S_A^*)-A(\widehat S_B)
    = \sum_{o\in O}\Delta_o-\sum_{p\in P}\Delta_p
    \le 2\varepsilon |O|
    \le 2B\varepsilon .
\]
\end{proof}

\begin{theorem}[Theorem A: marginal proxy regret]
\label{thm:ch6-a}
Under Assumptions~\ref{assum:ch6-a-binary}--\ref{assum:ch6-a-calibration},
let $S_B^*$ be an exhaustive maximiser of $G$ over size-$B$ schedules, and let
$\widehat S_B$ be top-$B$ under the proxy scores $\psi_t$. Then
\[
  G(S_B^*)-G(\widehat S_B)
  \le 2B\varepsilon + 2\eta_B .
\]
\end{theorem}

\begin{proof}
By approximate additivity, applied once at $S_B^*$ and once at
$\widehat S_B$,
\[
\begin{aligned}
  G(S_B^*)-G(\widehat S_B)
  &\le A(S_B^*)+\eta_B - \bigl(A(\widehat S_B)-\eta_B\bigr)\\
  &= A(S_B^*)-A(\widehat S_B)+2\eta_B .
\end{aligned}
\]
Let $S_A^*\in\argmax_{|S|=B}A(S)$. Since $S_A^*$ maximises $A$ over the same
feasible set,
\[
  A(S_B^*)\le A(S_A^*).
\]
Therefore
\[
  G(S_B^*)-G(\widehat S_B)
  \le A(S_A^*)-A(\widehat S_B)+2\eta_B .
\]
Lemma~\ref{lem:ch6-proxy-regret} bounds the additive proxy regret by
$2B\varepsilon$, giving the result.
\end{proof}

\paragraph{Interpretation.}
The bound has two terms. The calibration term $2B\varepsilon$ measures how
well the separable proxy orders the single-step marginal gains. The additivity
term $2\eta_B$ measures how much schedule value deviates from the sum of
single-step effects. If both terms are small relative to the available
headroom, marginal timing is justified.

The theorem is intentionally conservative. Its uniform $L^\infty$ assumptions
range over all schedules of size at most $B$ and all time steps. In a noisy
high-dimensional generation pipeline, these suprema can be too large to give a
non-vacuous numerical guarantee, even when the theorem is mathematically true.
For this reason Theorem A is the baseline, not the central empirical claim of
the thesis. It tells us what must be true for marginal signal ranking to work;
the experiments test whether this regime actually holds.

\section{\texorpdfstring{Diagnostics $A'$ and $A''$}{Diagnostics A-prime and A-double-prime}}
\label{sec:ch6-a-diagnostics}

The quantities called A$'$ and A$''$ in this thesis are diagnostics. They are
not additional regret theorems.

\paragraph{Diagnostic A$'$: additivity scale.}
Given a schedule distribution $D_B$ or a fixed candidate pool $C_B$, define
\[
  \eta^{\mathrm{typ}}_{B,D}
  := \E_{S\sim D_B}\abs{G(S)-A(S)}
\]
and, for a finite pool,
\[
  \sigma^{\mathrm{pool}}_{B}
  := \operatorname{sd}_{S\in C_B}\bigl(G(S)-A(S)\bigr).
\]
These measure the typical size of the additive residual on the sampled
schedules. They do not imply a uniform bound on
$\abs{G(S)-A(S)}$ at the selected top schedule.

\paragraph{Diagnostic A$''$: rankability.}
Given a schedule sample or candidate pool, define
\[
  R_B := \rho_{\mathrm{Spearman}}\bigl(A(S),G(S)\bigr),
\]
where the correlation is computed over schedules in the sample. This measures
whether the additive surrogate ranks schedules similarly to the true schedule
gain on that sample. A high global rank correlation is useful evidence, but it
does not by itself bound the regret of the optimiser-selected schedule.

The distinction matters because selected-schedule regret is an extremal
quantity. A diagnostic can be small on average or favourable in correlation
while still missing the schedules selected by an optimiser. To turn such
measurements into a theorem one needs a uniform or high-probability statement
over the relevant candidate pool. That is the role of Theorem~B$'$ below.

\section{Finite-Pool Selected-Schedule Form}
\label{sec:ch6-finite-pool-a}

The experimentally usable form of the marginal baseline is a finite-pool
statement. Let $C_B\subseteq \{S:|S|=B\}$ be a fixed candidate pool. Let
$S_C^*\in\argmax_{S\in C_B}G(S)$ be the pool oracle. Suppose that, on this
pool,
\[
  \eta_{B,C}:=\sup_{S\in C_B}\abs{G(S)-A(S)}
\]
is finite. Suppose also that an estimated additive surrogate $\widehat A$
satisfies, with probability at least $1-\delta$,
\[
  \sup_{S\in C_B}\abs{A(S)-\widehat A(S)}
  \le \varepsilon_{B,C,\delta}.
\]
If $\widehat S\in C_B$ approximately maximises $\widehat A$ over $C_B$ with
optimisation gap $\omega_B$, then Theorem~B$'$ with $Q:=A$ gives
\[
  G(S_C^*)-G(\widehat S)
  \le
  2\eta_{B,C}+2\varepsilon_{B,C,\delta}+\omega_B
\]
with probability at least $1-\delta$.

Against the exhaustive oracle $S_B^*$ one must add the pool-approximation term
\[
  \kappa_B := G(S_B^*)-G(S_C^*) \ge 0,
\]
giving
\[
  G(S_B^*)-G(\widehat S)
  \le
  \kappa_B+2\eta_{B,C}+2\varepsilon_{B,C,\delta}+\omega_B .
\]
The term $\kappa_B$ is not directly observable unless the pool is known to
contain the exhaustive oracle. In experiments it can only be discussed
relative to an MC-oracle or another explicit pool.

The candidate pool must be constructed without leakage. It may be sampled
randomly, generated from training seeds, or generated from a surrogate trained
without held-out true-$G$ values. It must not be selected using held-out
evaluation gains and then used to claim held-out regret control.

\section{Empirical Ranker-Class Limitation}
\label{sec:ch6-ranker-limitation}

The marginal baseline also yields a scoped limitation on a class of rankers.
This limitation has a formal part and an empirical part, and neither part is a
universal impossibility result.

\begin{proposition}[Time-only mean-profile limitation]
\label{prop:ch6-ranker-limitation}
Consider the seed-averaged additive problem
\[
  \bar A(S)=\sum_{t\in S}\bar\Delta_t,
  \qquad
  \bar\Delta_t := \E_i[\Delta_{i,t}].
\]
Among all time-only separable rankers, the largest achievable additive value is
obtained by selecting the $B$ largest entries of the mean profile
$\bar\Delta_t$.
\end{proposition}

\begin{proof}
A time-only ranker assigns each step $t$ a score depending only on $t$ (or on a
deterministic function of $t$, such as phase). It therefore chooses a fixed
size-$B$ subset of time indices, independent of the seed. The additive
objective for such a fixed subset is exactly $\bar A(S)$. By
Lemma~\ref{lem:ch6-topb-additive}, the maximiser of $\bar A(S)$ is any set of
the $B$ largest $\bar\Delta_t$.
\end{proof}

The empirical statement used in this thesis is narrower. On the ProSeCo-OWT
case study, the tested separable rankers do not recover the best-of-$N$
MC-oracle headroom, and the mean-$\bar\Delta_t$ envelope enters the
no-detectable-gain band by $B=8$ in the existing April 2026 results. This
supports a limitation of the tested separable ranker class on that backend.

The proposition and the empirical observation do not rule out
feature-conditioned separable rankers outside the tested class. They do not
rule out pairwise schedulers, online controllers, or search procedures using
true-$G$ feedback. They also do not rule out non-separable uses of PRISM or
other informed-refinement mechanisms. What is included in the marginal
baseline is PRISM-style quality mass $QM_t$ when it is reduced to a separable
per-step score $\psi_t$.

\section{Pairwise Interactions}
\label{sec:ch6-pairwise-interactions}

If marginal additivity is inadequate, the next structured hypothesis is that
the dominant non-additivity is pairwise. Define the operational second
difference
\[
  \xi_{t,t'}
  := G(\{t,t'\})-\Delta_t-\Delta_{t'},
  \qquad t\ne t' .
\]
This is not a mechanistic decomposition of the neural network. It is a
second-order finite difference of the schedule-value function. It aggregates
all ways in which applying the corrector at one step changes the value of
applying it at another step: direct interactions, altered downstream states,
stochastic sensitivity, and metric effects.

The pairwise surrogate is
\[
  Q(S)
  :=
  A(S)
  +
  \sum_{\substack{t<t'\\ t,t'\in S}}
  \xi_{t,t'} .
\]
This $Q(S)$ is a schedule-level surrogate. It is unrelated to the historical
raw key \texttt{Q\_t}, which denotes the quality-mass signal $QM_t$.

For schedules of size two, $Q(S)=G(S)$ by construction. For larger schedules,
$Q(S)$ ignores triple and higher-order residuals. For example, for
$S=\{t_1,t_2,t_3\}$ the residual
\[
  G(S)-Q(S)
\]
contains all effects that cannot be represented by singleton and pair terms.
Theorem B is useful only when these higher-order residuals are small on the
schedules being considered.

\section{Theorem B: Exact Pairwise Surrogate}
\label{sec:ch6-theorem-b-exact}

Theorem B is a generic surrogate-regret theorem. Its value in the thesis comes
from using it as a falsifiable diagnostic: if $Q$ approximates $G$ better than
$A$ on actual size-$B$ schedules, then interaction-aware scheduling is
justified.

\begin{assumption}[Pairwise surrogate approximation]
\label{assum:ch6-b-zeta}
There exists $\zeta_B\ge 0$ such that
\[
  \abs{G(S)-Q(S)}\le \zeta_B
  \qquad\text{for every } S \text{ with } |S|=B .
\]
\end{assumption}

\begin{theorem}[Theorem B: exact surrogate regret]
\label{thm:ch6-b-exact}
Let
\[
  S_Q^* \in \argmax_{|S|=B}Q(S).
\]
Suppose $\widehat S$ has surrogate optimisation gap $\omega_B$, meaning
\[
  Q(S_Q^*)-Q(\widehat S)\le \omega_B .
\]
Under Assumption~\ref{assum:ch6-b-zeta},
\[
  G(S_B^*)-G(\widehat S)
  \le 2\zeta_B+\omega_B .
\]
\end{theorem}

\begin{proof}
By the surrogate approximation assumption,
\[
  G(S_B^*) \le Q(S_B^*)+\zeta_B .
\]
Since $S_Q^*$ maximises $Q$ over size-$B$ schedules,
\[
  Q(S_B^*)\le Q(S_Q^*).
\]
Thus
\[
  G(S_B^*)\le Q(S_Q^*)+\zeta_B .
\]
For the selected schedule $\widehat S$, the same approximation gives
\[
  G(\widehat S)\ge Q(\widehat S)-\zeta_B .
\]
The optimisation gap gives $Q(\widehat S)\ge Q(S_Q^*)-\omega_B$. Therefore
\[
  G(\widehat S)
  \ge Q(S_Q^*)-\omega_B-\zeta_B .
\]
Subtracting the lower bound on $G(\widehat S)$ from the upper bound on
$G(S_B^*)$ yields
\[
  G(S_B^*)-G(\widehat S)
  \le 2\zeta_B+\omega_B .
\]
\end{proof}

\section{Theorem B with an Estimated Surrogate}
\label{sec:ch6-theorem-b-estimated}

In experiments $Q$ is not usually known. The pairwise interactions must be
estimated, giving an estimated surrogate $\widehat Q$.

\begin{assumption}[Estimated surrogate accuracy]
\label{assum:ch6-b-alpha}
There exists $\alpha_B\ge 0$ such that
\[
  \abs{Q(S)-\widehat Q(S)}
  \le \alpha_B
  \qquad\text{for every } S \text{ with } |S|=B .
\]
\end{assumption}

\begin{theorem}[Theorem B: estimated surrogate regret]
\label{thm:ch6-b-estimated}
Assume Assumptions~\ref{assum:ch6-b-zeta} and~\ref{assum:ch6-b-alpha}. Let
$\widehat S$ approximately maximise $\widehat Q$ with gap $\omega_B$:
\[
  \widehat Q(S_{\widehat Q}^*)-\widehat Q(\widehat S)\le \omega_B,
  \qquad
  S_{\widehat Q}^*\in\argmax_{|S|=B}\widehat Q(S).
\]
Then
\[
  G(S_B^*)-G(\widehat S)
  \le 2\zeta_B+2\alpha_B+\omega_B .
\]
\end{theorem}

\begin{proof}
It is enough to show that $\widehat S$ has $Q$-optimisation gap at most
$\omega_B+2\alpha_B$. Let $S_Q^*\in\argmax_{|S|=B}Q(S)$. Since
$S_{\widehat Q}^*$ maximises $\widehat Q$,
\[
  \widehat Q(S_{\widehat Q}^*)\ge \widehat Q(S_Q^*).
\]
Using the assumed optimisation gap,
\[
  \widehat Q(\widehat S)
  \ge \widehat Q(S_{\widehat Q}^*)-\omega_B
  \ge \widehat Q(S_Q^*)-\omega_B .
\]
By the uniform estimation bound,
\[
  Q(\widehat S)
  \ge \widehat Q(\widehat S)-\alpha_B
  \ge \widehat Q(S_Q^*)-\omega_B-\alpha_B
  \ge Q(S_Q^*)-\omega_B-2\alpha_B .
\]
Hence
\[
  Q(S_Q^*)-Q(\widehat S)\le \omega_B+2\alpha_B .
\]
Applying Theorem~\ref{thm:ch6-b-exact} with this optimisation gap gives
\[
  G(S_B^*)-G(\widehat S)
  \le 2\zeta_B+\omega_B+2\alpha_B .
\]
\end{proof}

The constant in front of $\alpha_B$ is $2$. One $\alpha_B$ is paid when
translating the estimated value of the selected schedule back to $Q$, and one
$\alpha_B$ is paid when comparing the estimated value of the $Q$-optimal
schedule to its true surrogate value.

\section{\texorpdfstring{Theorem $B'$}{Theorem B-prime}: Finite-Pool High-Probability Form}
\label{sec:ch6-theorem-b-prime}

Uniform control over all $\binom{T}{B}$ schedules is usually unrealistic.
The experimentally usable theorem is therefore a finite-pool,
high-probability statement.

\begin{theorem}[Theorem B$'$: finite-pool surrogate regret]
\label{thm:ch6-b-prime}
Fix a candidate pool
\[
  C_B\subseteq \{S\subseteq\{1,\ldots,T\}: |S|=B\}.
\]
Define the pool oracle
\[
  S_C^* \in \argmax_{S\in C_B}G(S),
\]
the pairwise approximation error on the pool
\[
  \zeta_{B,C}:=\sup_{S\in C_B}\abs{G(S)-Q(S)},
\]
and suppose that with probability at least $1-\delta$,
\[
  \sup_{S\in C_B}\abs{Q(S)-\widehat Q(S)}
  \le \alpha_{B,\delta}.
\]
Let $\widehat S\in C_B$ approximately maximise $\widehat Q$ over $C_B$ with
gap $\omega_B$. Then, with probability at least $1-\delta$,
\[
  G(S_C^*)-G(\widehat S)
  \le 2\zeta_{B,C}+2\alpha_{B,\delta}+\omega_B .
\]
If $S_B^*$ denotes the exhaustive oracle and
\[
  \kappa_B:=G(S_B^*)-G(S_C^*),
\]
then, on the same event,
\[
  G(S_B^*)-G(\widehat S)
  \le
  \kappa_B
  +2\zeta_{B,C}
  +2\alpha_{B,\delta}
  +\omega_B .
\]
\end{theorem}

\begin{proof}
The proof of Theorem~\ref{thm:ch6-b-estimated} applies verbatim after
restricting every maximisation and supremum to $C_B$. This gives the first
inequality against $S_C^*$. The second follows by adding and subtracting
$G(S_C^*)$:
\[
  G(S_B^*)-G(\widehat S)
  =
  \bigl(G(S_B^*)-G(S_C^*)\bigr)
  +
  \bigl(G(S_C^*)-G(\widehat S)\bigr).
\]
The first term is $\kappa_B$ by definition.
\end{proof}

The no-leakage requirement is part of the theorem's use, not a technicality.
The candidate pool $C_B$ must be fixed independently of held-out evaluation
values of $G$. A random MC pool is admissible. A pool generated by a surrogate
trained on training seeds is admissible when evaluated on held-out seeds. A
pool chosen because its schedules have high held-out true-$G$ values is not
admissible for a held-out regret claim.

For Phase 1 schedule-level validation, $Q(S)$ must be computable for the
schedules in $C_B$. This means either every pair inside each schedule has a
measured $\xi_{t,t'}$, or $\xi_{t,t'}$ is supplied by a pre-registered
estimator fitted only on training pairs and training seeds. Missing
interactions must not be silently filled using held-out $G$ values. If neither
measured pairs nor a no-leakage estimator covers a schedule, that schedule is
not eligible for the $Q(S)\approx G(S)$ validation pool.

\section{Levels of Analysis}
\label{sec:ch6-levels}

The same symbols can be instantiated at three different levels. The level
determines what can be claimed.

\paragraph{Level 1: seed-wise diagnostic.}
For a fixed seed $i$, common random numbers define a seed-wise gain
$G_i(S)$, marginals $\Delta_{i,t}$, interactions $\xi_{i,t,t'}$, and
surrogate $Q_i(S)$. These quantities are diagnostic. They require
counterfactual evaluations on the same trajectory and are not by themselves
deployable.

\paragraph{Level 2: population or global schedule.}
Averaging over paired seeds gives
\[
  \bar G(S)=\E_i[G_i(S)],
  \qquad
  \bar\Delta_t=\E_i[\Delta_{i,t}],
  \qquad
  \bar\xi_{t,t'}=\E_i[\xi_{i,t,t'}].
\]
Optimising a population surrogate produces one global schedule that is good on
average. It does not adapt to features of an individual trajectory.

\paragraph{Level 3: feature-conditioned scheduler.}
A deployable scheduler must construct $\widehat Q_i(S)$, or an equivalent
decision rule, from observable features of the current trajectory, trained on
other seeds and evaluated on held-out seeds. It cannot use true
counterfactual $G_i(S)$ at inference time. Only Level 3 supports a deployable
scheduler claim.

This hierarchy prevents overclaiming. Level 1 can show that pairwise structure
exists. Level 2 can show that a global schedule or population pairwise
surrogate is useful. Level 3 is required to claim that the framework yields a
practical trajectory-conditioned scheduler.

\section{Diagnostic Framework C}
\label{sec:ch6-framework-c}

Diagnostic Framework C is a protocol, not a theorem. Given a fixed tuple
(model, corrector, $F$, $B$), it classifies the empirical situation into
regimes by measuring headroom, marginal rankability, interaction strength,
pairwise predictability, and closure of the available headroom.

Fix a paired seed set and a pre-specified schedule pool. Define
\[
  U_B^{\mathrm{MC},N}
  :=
  G(S_B^{\mathrm{MC},N})-G(S_B^{\mathrm{uniform}}),
\]
the best-of-$N$ MC-oracle headroom over a uniform-placement baseline. Define
\[
  U_B^{\mathrm{pool}}
  :=
  G(S_C^*)-G(S_B^{\mathrm{uniform}}),
\]
the pool-oracle headroom on $C_B$. The unobservable exhaustive-oracle headroom
may be denoted $U_B^*$ in theory, but it is not reported as an empirical
quantity.

The marginal rankability diagnostic is
\[
  R_B := \rho_{\mathrm{Spearman}}\bigl(A(S),G(S)\bigr).
\]
An interaction-strength diagnostic is
\[
  I_B :=
  \frac{
    \operatorname{sd}\bigl(G(S)-A(S)\bigr)
  }{
    \operatorname{sd}\bigl(A(S)\bigr)+\delta_0
  },
\]
where $\delta_0>0$ is a small stabilising constant. The pairwise
predictability diagnostic is
\[
  P_B := \rho_{\mathrm{Spearman}}\bigl(Q(S),G(S)\bigr).
\]
For a method producing a schedule $\widehat S$, define a closure ratio against
an MC pool by
\[
  C_B^{\mathrm{MC},N}(\widehat S)
  :=
  \frac{G(\widehat S)-G(S_B^{\mathrm{uniform}})}
       {G(S_B^{\mathrm{MC},N})-G(S_B^{\mathrm{uniform}})}
\]
whenever the denominator is statistically distinguishable from zero.

The intended regimes are:
\begin{description}[leftmargin=2.5cm,style=nextline]
  \item[No-op.]
  The corrector-placement budget has no detectable positive headroom:
  $U_B^{\mathrm{MC},N}$ and $U_B^{\mathrm{pool}}$ are indistinguishable from
  zero. Timing is not meaningful at this budget.

  \item[Marginal/rankable.]
  Headroom exists, $R_B$ is high, interaction strength is low, and tested
  separable rankers recover most of the available pool or MC headroom. This is
  the regime where Theorem A is the right explanation.

  \item[Interaction-driven.]
  Headroom exists, additive rankability is weak or moderate, pairwise
  predictability improves over additive predictability, and
  $\zeta_{B,C}<\eta_{B,C}$ and/or $P_B>R_B$ with uncertainty accounted for.
  This is the regime where Theorem B and B$'$ justify pairwise scheduling.

  \item[Higher-order/chaotic.]
  Headroom exists, but neither $A(S)$ nor $Q(S)$ predicts $G(S)$ well on
  actual size-$B$ schedules. True-$G$ search may still find good schedules, but
  no compact additive or pairwise surrogate has been validated.

  \item[Online-decision-like.]
  Offline structure exists and a compact observable state supports a
  held-out, budget-aware policy that beats non-budget-aware baselines. This
  regime requires Level 3 evidence.
\end{description}

Classification is statistical. Statements such as $P_B>R_B$ or
$U_B^{\mathrm{MC},N}>0$ mean that the corresponding paired confidence interval
supports the comparison. Pair heatmaps and phase-pair summaries are useful
diagnostics, but they do not by themselves validate Theorem B. The validation
must occur at the schedule level: compare $A(S)$, $Q(S)$, and $G(S)$ on actual
size-$B$ schedules in a no-leakage candidate pool.

\section{Appendix-Grade Directions}
\label{sec:ch6-appendix-grade}

Two mathematical objects are useful for orientation but are not load-bearing in
the current thesis.

\paragraph{Theorem D: online controller abstraction.}
One can formulate corrector timing as a finite-horizon budgeted control
problem. Let $z_t$ be an observable state, $b_t$ the remaining placement
budget, and $a_t\in\{0,1\}$ the decision to apply the corrector. The value
function satisfies a Bellman recursion
\[
  V_t(z,b)
  =
  \max_{a\in\{0,1\},\,a\le b}
  \E\!\left[
    V_{t+1}(z_{t+1},b-a)
    \mid z_t=z, a_t=a
  \right],
\]
with terminal reward given by the final quality. If an approximate value
function $\widehat V_t$ is uniformly within $\delta$ of $V_t$ on all reachable
states, the standard approximate dynamic programming argument gives a
$2T\delta$ performance-loss bound for the greedy policy. This is generic. It
becomes thesis-central only if a compact observed state yields a positive
held-out scheduler, which has not yet been shown.

\paragraph{Lemma E: burn-in under clipped quality.}
Early denoising steps may have small or empty revisable sets, suggesting that
corrector placements there have little value. A clean deterministic lemma can
be stated for a clipped or bounded-Lipschitz surrogate $F_C$: if
\[
  \abs{F_C(y)-F_C(y')}\le L_F d_H(y,y'),
\]
then the gain from a correction acting on a small revisable set is bounded by a
constant proportional to $\abs{R_t}$. This does not apply unconditionally to
the experimental score $F=-$GPT-2 NLL, which need not be Lipschitz in Hamming
distance without clipping. In the main empirical pipeline, burn-in exclusions
should therefore be justified by measured $R_t$ and measured $\Delta_t$, not by
an unconditional Lipschitz claim.

\section{Conclusion}
\label{sec:ch6-conclusion}

The chapter gives a linear framework for fixed-budget corrector timing.
Theorem~\ref{thm:ch6-a} says when marginal ranking should work: the schedule
gain must be approximately additive, and observable signals must calibrate the
single-step marginal gains. Diagnostics A$'$ and A$''$ measure additivity scale
and rankability, but they are not regret refinements. The finite-pool form,
obtained as Theorem~B$'$ with $Q:=A$, is the safe selected-schedule statement
used in experiments.

When marginal ranking is insufficient, the pairwise framework takes over.
The interaction term $\xi_{t,t'}$ defines a second-order schedule surrogate
$Q(S)$. Theorem~\ref{thm:ch6-b-exact} proves the exact surrogate regret bound,
Theorem~\ref{thm:ch6-b-estimated} adds estimated-surrogate error with the
constant $2\alpha_B$, and Theorem~\ref{thm:ch6-b-prime} gives the finite-pool,
high-probability, no-leakage form that can be tested experimentally.

Diagnostic Framework C then classifies the observed regime: no-op,
marginal/rankable, interaction-driven, higher-order/chaotic, or
online-decision-like. The experiments in the rest of the thesis ask which of
these regimes ProSeCo-OWT occupies. If the marginal assumptions hold, Theorem A
explains why simple rankers work. If pairwise schedule-level validation holds,
Theorem B/B$'$ explains why interaction-aware scheduling is justified. If both
fail while true-$G$ search still finds headroom, the framework still provides a
mathematically precise negative result: the tested backend lies outside the
validated marginal and pairwise regimes, and the contribution becomes a
diagnostic classification rather than a deployable scheduler.
