\chapter{Experiments}
\label{ch:experiments}

This chapter is intentionally a scaffold. It should be written only after
Phase 0 re-confirms the real ProSeCo-OWT backend path and the old qualitative
pattern. Until then, prior results are background for planning, not final
thesis evidence.

\section{Experimental Setup}
\label{sec:exp-setup}

To be written after Phase 0. The setup should state the fixed predictor
schedule, fixed informed corrector, placement budget $B$, extra-NFE budget
$B_{\mathrm{NFE}} = c_{\mathrm{corr}}B$, paired common-random-number protocol,
quality functional $F$, evidence tiers, and bootstrap procedure.

\section{Phase 0 Reproducibility}
\label{sec:exp-phase0}

To be written after the checkpoint-backed PF1--PF8 checks and K=3 smoke.
This section should report whether the current code path reproduces the old
qualitative ProSeCo-OWT keys before any new experiment is trusted.

\section{ProSeCo-OWT Case Study}
\label{sec:exp-proseco-owt}

To be written after Phase 0. The main body should report the ProSeCo-OWT
case study with current scoped language: tested separable rankers do not
recover MC-oracle headroom under the verified protocol, while CD-G and BS-AG
are search-based existence results using true schedule-gain feedback.

\section{Interaction Diagnostics}
\label{sec:exp-interaction-diagnostics}

To be written after Phase 1. This section should report schedule-level
validation of $Q(S)$ against $G(S)$ on no-leakage size-$B$ candidate pools,
not only pairwise heatmaps. It should state whether Theorem B/B$'$ is useful
on the verified pool and at which level of analysis.

\section{Discussion}
\label{sec:exp-discussion}

To be written after Phase 1 / Phase 2 gates. The discussion should preserve
the single-backbone scope, distinguish MC-oracle and pool-oracle headroom from
the unobservable exhaustive oracle, and avoid deployable scheduler claims
unless a Level-3 feature-conditioned scheduler succeeds on held-out seeds.
